---
title: "The tidyverse: dplyr, ggplot2, and friends"
output: html_document
---

This lesson covers packages primarily by Hadley Wickham for tidying data and then working with it in tidy form, collectively known as the "tidyverse".

The packages we are using in this lesson are all from CRAN, so we can install them with `install.packages`. Don't run this if you are using our biotraining server, the packages are already installed!

```{r eval=FALSE}
# install.packages(c(
#     "tidyverse",
#     "viridis",
#     "broom"
# ))
```

```{r warning=FALSE, message=FALSE}
library(tidyverse) # Load all "tidyverse" libraries.
# OR
# library(readr)   # Read tabular data.
# library(tidyr)   # Data frame tidying functions.
# library(dplyr)   # General data frame manipulation.
# library(ggplot2) # Flexible plotting.

library(viridis)   # Viridis color scale.
```

These packages usually have useful documentation in the form of "vignettes". These are readable on the CRAN website, or within R:

```{r eval=FALSE}
vignette()
vignette(package="dplyr")
vignette("dplyr", package="dplyr")
```


Let's continue our examination of the FastQC output. If you're starting fresh for this lesson, you can load the necessary data frame with:

```{r}
bigtab <- read_csv("r-more-files/fastqc.csv")
```

# ggplot2 revisited

We [saw ggplot2 in the introductory R day](https://monashdatafluency.github.io/r-intro-2/plotting-with-ggplot2.html). Recall that we could assign columns of a data frame to aesthetics--x and y position, color, etc--and then add "geom"s to draw the data.

With `ggplot2` we can easily view the whole data set.

```{r}
ggplot(bigtab, aes(x=file,y=test,color=grade)) + 
    geom_point()
```

With categorical data on the x and y axes, a better geom to use is `geom_tile`.

```{r}
ggplot(bigtab, aes(x=file,y=test,fill=grade)) + 
    geom_tile()
```


## Publication quality images

`ggplot2` offers a very wide variety of ways to adjust a plot. For categorical aesthetics, usually the first step is ensuring the relevant column is a factor with a meaningful level order.

```{r}
y_order <- sort(unique(bigtab$test), decreasing=T)  # y axis plots from bottom to top, so reverse
bigtab$test <- factor(bigtab$test, levels=y_order)

x_order <- unique(bigtab$file)
bigtab$file <- factor(bigtab$file, levels=x_order)

# Only necessary if not continuing from previous lesson on programming!
color_order <- c("FAIL", "WARN", "PASS")
bigtab$grade <- factor(bigtab$grade, levels=color_order)

myplot <- ggplot(bigtab, aes(x=file, y=test, fill=grade)) + 
    geom_tile(color="black", size=0.5) +           # Black border on tiles
    scale_fill_manual(                             # Colors, as color hex codes
        values=c("#ee0000","#ffee00","#00aa00")) +
    labs(x="", y="", fill="") +                    # Remove axis labels
    coord_fixed() +                                # Square tiles
    theme_minimal() +                              # Minimal theme, no grey background
    theme(panel.grid=element_blank(),              # No underlying grid lines
          axis.text.x=element_text(                # Vertical text on x axis
              angle=90,vjust=0.5,hjust=0))              
myplot
```

Plots can be saved with ``ggsave``. Width and height are given in inches, and an image resolution in Dots Per Inch should also be given. The width and height will determine the relative size of text and graphics, so increasing the resolution is best done by adjusting the DPI. Compare the files produced by:

```{r, eval=FALSE}
ggsave("plot1.png", myplot, width=5,  height=5,  dpi=600)
ggsave("plot2.png", myplot, width=10, height=10, dpi=300)
```

It may be necessary to edit a plot further in a program such as Inkscape or Adobe Illustrator. To allow this, the image needs to be saved in a "vector" format such as SVG, EPS, or PDF. In this case, the DPI argument isn't needed.


# dplyr

`dplyr` gives us a collection of convenient "verbs" for manipulating data frames. The name is a contraction of "Data frame apPLYeR", as some of these verbs have a similar flavour to `apply`/`lapply`/`tapply`/etc.

Each verb takes a data frame as input and returns a modified version of it. The idea is that complex operations can be performed by stringing together a series of simpler operations in a pipeline.

```
# input         +--------+        +--------+        +--------+      result
#  data   %>%   |  verb  |  %>%   |  verb  |  %>%   |  verb  |  ->   data
#   frame       +--------+        +--------+        +--------+        frame
```


## tibbles

```{r}
bigtab
```

As we used `readr` to read the data, `bigtab` is a "tibble", which is the Tidyverse's improved data frame. The `dplyr` verbs also all return their results as tibbles. You can also create tibbles explicitly with the `tibble` function (or `data_frame`). One convenient feature is that it only shows a few rows of the data frame when you print it. If you do want to see the whole table, you can use `as.data.frame`, or use the `View` function to view it in a tab.

```{r eval=FALSE}
as.data.frame(bigtab)
View(bigtab)
```

The `n` and `width` arguments to `print` can also be used to print more rows or columns respectively.

```{r eval=FALSE}
print(bigtab, n=100, width=1000)
```


## filter

Say we want to know all the tests that failed. `filter` provides a convenient way to get these.

```{r}
filter(bigtab, grade == "FAIL")
```

Something is magic here: we do not have `grade` in our environment, only within the data frame. Arguments to `filter` can use any column of the data frame as a variable. `dplyr` uses this idea heavily, arguments to `dplyr` verbs often behave in a special way.


## arrange

Rather than filtering, we might instead want to sort the data so the most important rows are at the top. `arrange` sorts a data frame by one or more columns.

```{r}
arrange(bigtab, grade)

# desc( ) can be used to reverse the sort order
arrange(bigtab, desc(grade))
```


## select

dplyr's `select` function is for subsetting, renaming, and reordering *columns*. Old columns can be referred to by name or by number.

```{r}
select(bigtab, test,grade) 
select(bigtab, 2,1)
select(bigtab, foo=file, bar=test, baz=grade)
```

You can also remove specific columns like this:

```{r}
select(bigtab, -file)
```

The magic here is that column variables refer to their column number. `select` subsets columns rather like subsetting vectors with `[ ]`. So `-file` becomes `-1`, and negative numbers remove items. We can also refer to ranges of columns, such as `grade:file`, which would become `1:3`, which is `c(1,2,3)`.


## Joins

Say we want to convert PASS/WARN/FAIL into a numeric score, so we can produce some numerical summaries. The scoring scheme will be:

```{r}
fwp <- c("FAIL","WARN","PASS")
scoring <- tibble(grade=factor(fwp,levels=fwp), score=c(0,0.5,1))

# Or: 
# scoring <- data.frame(grade=factor(fwp,levels=fwp), score=c(0,0.5,1))

scoring
```

In the introduction to R day, we saw the `merge` function. dplyr has its own version of this in the form of several functions: `left_join`, `right_join`, `inner_join`, `full_join`, `anti_join`.

The difference between these functions is what happens when there is a row in one data frame without a corresponding row in the other data frame. `inner_join` discards such rows. `full_join` always keeps them, filling in missing data with NA. `left_join` always keeps rows from the first data frame. `right_join` always keeps rows from the second data frame. `anti_join` is a bit different, it gives you rows from the first data frame that *aren't* in the second data frame.

Often we use a join to augment a data frame with some extra information. `left_join` is a good default choice for this as it will never delete rows from the data frame that is being augmented.

```{r}
scoretab <- left_join(bigtab, scoring, by="grade")
scoretab
```

"grade" is here acting as the *key*, allowing corresponding rows in the data frames to be joined together.

One important thing that all the join functions do: if multiple rows have the same key in either data frame, all ways of combining the two sets of rows will be included in the result. So, here, rows from the scoring data frame have been copied many times in the output. 

Joins are the key tool for *integrating* different types of data, based on some common key such as gene id.

<b>See also:</b> `match` and `merge` in base R.


### Challenge {.challenge}

Using the newly created `score` column:

1. Filter `scoretab` to get only "WARN" or "FAIL" grades. 

2. Sort the result so that "FAIL"s are at the top.


## mutate

`mutate` lets us add or overwrite columns by computing a new value for them.

```{r}
mutate(scoretab, doublescore = score*2)
```

Even though this is called "mutate", it is not literally modifying the input. Rather it is producing a copy of the data frame that has the modifiction.

This is equivalent to:

```{r eval=FALSE}
scoretab2 <- scoretab
scoretab2$doublescore <- scoretab2$score * 2
```


## summarize

`summarize` lets us compute summaries of data.

```{r}
summarize(scoretab, total=sum(score))
```

We really want to summarize the data grouped by file, so we can see if there are any particularly bad files. This is achieved using the `group_by` "adjective". `group_by` gives the data frame a grouping using one or more columns, which modifies the subsequent call to `summarize`.

```{r}
group_by(scoretab, file)

summarize(group_by(scoretab, file), average_score=mean(score))
```

The special function `n()` can be used within `summarize` to get the number of rows. (This also works in `mutate`, but is most useful in `summarize`.)

```{r}
summarize(group_by(scoretab, grade), count=n())
```

So `summarize` lets us do the things we do with base R using `table` and `tapply`, but staying tidy by using data frames.

<b>Tip:</b> `group_by` also affects other verbs, such as mutate. This is more advanced than we will be going into today. Grouping can be removed with `ungroup`.

See also functions for common uses of summarize: `count`, `distinct`.


# The pipe %>%

We often want to string together a series of `dplyr` functions. This is achieved using `dplyr`'s pipe operator, `%>%`. This takes the value on the left, and passes it as the first argument to the function call on the right. So the previous summarization could be written:

```{r eval=FALSE}
scoretab %>% group_by(grade) %>% summarize(count=n())
```

`%>%` isn't limited to `dplyr` functions. It's an alternative way of writing any R code.

```{r}
rep(paste("hello", "world"), 5)

"hello" %>% paste("world") %>% rep(5)
```

The idea of adding geoms in `ggplot2` is rather like the `dplyr` pipe. `ggplot2` predates `dplyr`, and Hadley Wickham has had a progression of ideas. It will probably be possible to use `%>%` instead of `+` in some successor to `ggplot2` in the not too distant future.



# tidyr

`tidyr` is the Tidyverse package for getting data frames to tidy. Recall that in a tidy data frame:

* each row is a unit of observation
* each column is a single piece of information

I'm not convinced `tidyr` is a great package, but the *ideas* behind it are very important. `dplyr` uses magic arguments to produces very beautiful R code, but in `tidyr` it gets a bit confusing. Anyway, let's work through a toy example:

```{r}
untidy <- read_csv(
    "country,     male-young, male-old, female-young, female-old
     Australia,            1,        2,            3,          4
     New Zealand,          5,        6,            7,          8")
untidy
```

The first problem is that rows are not distinct units of observation, there are actually four observations per row. This is fixed using `gather`.

```{r}
gathered <- gather(untidy, key=group, value=cases, -country)
gathered
```

We give a data frame, a name for the key and value columns to be created. Remaining arguments behave like `select`, here we are saying that the country column shouldn't be gathered.

In the introductory R day we saw `melt` from the `reshape2` package, which did a similar thing to matrices.

`spread` is the opposite operation, spreading a column into many columns, which generally makes data less tidy but is sometimes necessary. A common application is producing a scatter plot, where the x and y axes need to be two different columns even though they measure the same type of thing. Data may also be easier to look at in a table in spread form.

```{r}
spread(bigtab, key=file, value=grade)
```

In our toy example, we have a further problem that the "group" column contains two pieces of data. This can be fixed with `separate`. By default `separate` splits on any non-alphanumeric characters, but different separator characters can be specified.

```{r}
separate(gathered, col=group, into=c("gender","age"))
```

All of this would typically be written as a single pipline:

```{r}
tidied <- untidy %>%
    gather(key=group, value=cases, -country) %>%
    separate(group, into=c("gender","age"))
```

<b>Advanced:</b> `tidyr` has a number of other useful data frame manipulations. For completeness, we mention `nest`. This creates a *list column* in a tibble, the list column containing one tibble per row. Yo, Hadley put tibbles in your tibble so you can dplyr while you dplyr. The inverse operation is `unnest`. `unnest` is rather like the `bind_rows` function we saw earlier. `nest` and `unnest` go well with creating or modifying list columns in a `mutate` with `lapply` (or the `purrr` package).

```{r eval=FALSE}
# Advanced
nested <- nest(gathered, -country)
nested
nested$data
unnest(nested)
```


### Challenge {.challenge}

You receive data on a set of points. The points are in two dimensions (`dim`), and each point has x and y coordinates. Unfortunately it looks like this:

```{r}
df <- read_csv(
    "dim, A_1, A_2, B_1, B_2, B_3, B_4, B_5
     x,   2,   4,   1,   2,   3,   4,   5
     y,   4,   4,   2,   1,   1,   1,   2")
```

1. Tidy the data by *gathering* all of the columns except `dim`. What what does each row now represent?

2. We want to plot the points as a scatter-plot, using either `plot` or `ggplot`. *Spread* the gathered data so that this is possible. Now what do the rows represent?

3. What other tidying operation could be applied to this data?


# An RNA-Seq example

We will now put this all together by looking at a small-scale gene expression data set. To simplify somewhat: The data was generated from a multiplex PCR targetting 44 genes. The PCR product is quantified by counting reads from high throughput sequencing. The experiment is of yeast released synchronously into cell cycling, so that we can see which genes are active at different points in the cycle. Several strains of yeast have been used, a wildtype and several gene knock-downs. Each has nine time points, covering two cell cycles (two hours). 

This is much smaller than a typical RNA-Seq dataset. We are only looking at 44 genes, rather than all of the thousands of genes in the yeast genome.

Recall the three activities involved in data analysis. This dataset requires all of them.

<img src="../figures/data-cycle.svg" height="400">


## Tidying

<b>Hint:</b> You can select from the top of a pipeline to partway through and press Ctrl-Enter or Command-Enter to see the value part-way through the pipeline.

```{r}
# Use readr's read_csv function to load the file
counts_untidy <- read_csv("r-more-files/read-counts.csv")

counts <- counts_untidy %>%
    gather(key=sample, value=count, -Feature, factor_key=TRUE) %>%
    separate(sample, sep=":", into=c("strain","time"), convert=TRUE, remove=FALSE) %>%
    mutate(
        strain = factor(strain, levels=c("WT","SET1","RRP6","SET1-RRP6"))
    ) %>%
    filter(time >= 2) %>%
    select(sample, strain, time, gene=Feature, count)

summary(counts)
```

Time point 1 was omitted because it isn't actually part of the time series, it's from cells in an unsynchronized state. If we wanted to be even tidier, the remaining time points could be recoded with the actual time within the experiment -- they are at 15 minute intervals.


## Transformation and normalization

```{r}
ggplot(counts, aes(x=sample, y=count)) + 
    geom_boxplot() + 
    coord_flip()
```

We immediately see from a [Tukey](https://www.amazon.com/Exploratory-Data-Analysis-John-Tukey/dp/0201076160/ref=sr_1_1?s=books&ie=UTF8&qid=1469077563&sr=1-1) box plot that the data is far from normal -- there are outliers many box-widths to the right of the box. Perhaps a log transformation is in order.

```{r}
ggplot(counts, aes(x=sample, y=log2(count))) + 
    geom_boxplot() + 
    coord_flip()

ggplot(counts, aes(x=log2(count), group=sample)) + 
    geom_line(stat="density")
```

Log transformation has greatly improved things, although now there are more outliers left of the whiskers, and some zeros we will need to be careful of. We also see one of the samples has less reads than the others.

The gene SRP68 was included as a control housekeeping gene. Its expression level should be the same in all samples. We will divide counts in each sample by the count for SRP68 to correct for library size, then log-transform the data. We add a small constant when log-transforming so that zeros do not become negative infinity.

```{r}
normalizer <- counts %>%
    filter(gene == "SRP68") %>%
    select(sample, norm=count)

moderation <- 0.5/mean(normalizer$norm)

counts_norm <- counts %>%
    left_join(normalizer, by="sample") %>%
    mutate(
        norm_count = count/norm, 
        log_norm_count = log2(norm_count+moderation)
    )

ggplot(counts_norm, aes(x=sample, y=log_norm_count)) + 
    geom_boxplot() + 
    coord_flip()
```

**Note:** Here all of the genes are potentially differentially expressed. For a full RNA-Seq data set we would probably assume most genes aren't differentially expressed, so rather than nominating a normalizing gene we would normalize by total library size with, typically with a slight adjustment using the "TMM" method as calculated by `calcNormFactors` in `edgeR`:

```{r eval=FALSE}
# For a full sized RNA-Seq dataset:
library(edgeR)
mat <- counts_untidy %>% select(-Feature) %>% as.matrix
adjusted_lib_sizes <- colSums(mat) * calcNormFactors(mat)
normalizer_by_tmm <- tibble(sample=names(adjusted_lib_sizes), norm=adjusted_lib_sizes)
```

## Visualization

The data is now suitably transformed and normalized for visualization. `ggplot2` gives us a lot of freedom to juxtapose information from different columns.

### Challenge {.challenge}

1. Get all the rows in `counts_norm` relating to the histone gene "HHT1".

2. Plot this data with ggplot2. Use `time` as the x-axis, `log_norm_count` as the y-axis, and color the data by `strain`. Try using geoms: `geom_point()`, `geom_line()`.

```{r eval=FALSE}
ggplot( ... , aes(x= ... , y= ... , color= ... )) + ...
```

Extensions: 

Compare plots of `log_norm_count`, `norm_count`, and `count`.

Experiment with other geoms and other ways to assign columns to aesthetics.


### Whole dataset visualization

```{r fig.width=8, fig.height=8}
ggplot(counts_norm, aes(x=sample, y=gene, fill=log_norm_count)) + 
    geom_tile() + 
    scale_fill_viridis() + 
    theme_minimal() +
    theme(axis.text.x=element_text(           # Vertical text on x axis
              angle=90,vjust=0.5,hjust=1))              
```

```{r fig.width=8, fig.height=8}
ggplot(counts_norm, aes(x=time, y=gene, fill=log_norm_count)) + 
    geom_tile() + 
    facet_grid(~ strain) + 
    scale_fill_viridis() + 
    theme_minimal()
```

```{r fig.width=10, fig.height=8}
ggplot(counts_norm, aes(x=time, y=strain, fill=log_norm_count)) + 
    geom_tile() + 
    facet_wrap(~ gene) + 
    scale_fill_viridis() + 
    theme_minimal()
```

```{r fig.width=10, fig.height=8}
ggplot(counts_norm, aes(x=time, y=log_norm_count, color=strain, group=strain)) + 
    geom_line() + 
    facet_wrap(~ gene, scale="free")
```


## Exercises {.challenge}

1. Which are the three most variable genes?

Hint: `intToUtf8(utf8ToInt("xvh#jurxsbe|/#vxppdul}h/#vg/#dqg#duudqjh")-3)`

2. Different genes have different average expression levels, but what we are interested in is how they change over time. Further normalize the data by subtracting the average for each gene from `log_norm_count`.


# Appendix: Tidy linear modelling

The idea of this section is to give a taste of what is possible, so that you know what topics to read up on or ask an expert for help with if this is the sort of analysis you need. It will also provide a little context for advanced forms of RNA-Seq differential expression analysis.

```{r}
sut476 <- counts_norm %>% filter(gene=="SUT476") 
sut476_wt <- sut476 %>% filter(strain=="WT")

ggplot(sut476_wt,aes(x=time,y=log_norm_count)) +
    geom_point() +
    geom_smooth(method="lm")
```

`ggplot2` includes a handy geom for fitting a curve or line to data, but what exactly is it doing? What if we want to work with that fitted curve ourselves? `ggplot2` can use various curve fitting methods, the simplest of which is a linear model with "lm".

```{r}
model <- lm(log_norm_count ~ time, data=sut476_wt)
model
# Note: The p-values listed by summary aren't always meaningful.
# It makes no sense to remove the intercept term but retain the time term.
# Significance testing can better be done with anova() and the multcomp package.
summary(model)
model.matrix(log_norm_count ~ time, data=sut476_wt)
```

A model can be tested against a simpler null model using `anova`. Is the fit sufficiently better to justify the extra complexity?

```{r}
null_model <- lm(log_norm_count ~ 1, data=sut476_wt)
anova(null_model, model)
```

A linear trend is not supported, in comparison to the null hypothesis that expression is constant, by this data. Also useful for hypothesis testing, the `multcomp` package can be used to test comparisons of interest such as a set of pairwise comparisons. This is similar to testing of contrasts in packages for differential expression analysis such as `limma`, `edgeR`, or `DESeq2`.

`augment` from the `broom` package gives various diagnostic statistics.

* .fitted = Model prediction.
* .residual = Actual value minus .fitted.
* .hat = The leverage -- how important the observation is in the design of the experiment. This can be considered before performing the experiment!
*  .cooksd = Cook's Distance -- how much actual influence the observation had on the model. Combines information from .residual and .hat.

```{r}
library(broom)

augment(model, sut476_wt) %>% head

# augment() can also be used to produce predictions for new inputs
augment(model, newdata=sut476_wt[1:5,])

# Let's look at the model fit
augment(model, sut476_wt) %>%
ggplot(aes(x=time)) + 
    geom_point(aes(y=log_norm_count)) +
    geom_line(aes(y=.fitted))
```

We see the initial timepoint in the wildtype strain is wildly off the linear trend. Possibly early time points need to be removed, or (more likely!) a linear trend is not a good model of this data.

Linear models can be much more complicated than this. They can include multiple explanatory variables, including categorical variables, and interactions between them, and even polynomials.

```{r}
model2 <- lm(log_norm_count ~ strain*poly(time,3), data=sut476)
model2

augment(model2, sut476) %>%
ggplot(aes(x=time, color=strain, group=strain)) + 
    geom_point(aes(y=log_norm_count), alpha=0.5) +
    geom_line(aes(y=.fitted))
```


```{r}
sessionInfo()
```



